---
title: "Joining shapfiles with csvs for use with QGis"
author: "Billy Ehrenberg"
date: "26 June 2017"
output: html_document
---

```{r setup, include=FALSE}
packages.needed <- c(
"dplyr",
"rgdal",
"rgeos",
"rmarkdown")

packages.new <- packages.needed[!(packages.needed %in% installed.packages()[,"Package"])]
if(length(packages.new)) install.packages(packages.new)
lapply(packages.needed, require, character.only = TRUE)

rm(list=ls())
```
## Getting started

This process will allow you to read a shapefile into R and then join it with a csv, before saving it as a new shapefile, meaning you can read it into QGis as one dataset containing polygons and data.

If you want yojust run the code, you can change the variables in the join_csv_with_shapefile.R script and it should work straight out of the box. If you want to understand the process, keep reading.

You should create a new folder to work in, so you can keep all your files in one place. The best way to do this is to start a new R project. That's easy:

<br>1) Go to file</br>
<br>2) Choose new project</br>
<br>3) Decide if you want to create a new folder to house your project or use an existing one</br>
<br>4) Choose the preferred project option</br>
<br>5) Click browse to choose where you want to save it</br>

When you create the new project, R will automatically know you are working in that folder, and will have easy access to all the files in it.

## What you'll need in your folder

A shapefile and a csv. 

They need to have a common column (a column where the names or codes are the same in the csv as in the shapfile). This common column should have the same name in both. If it doesn't to start with, that's easy to fix.

Shapefiles are kept in folders with several different files in them. That's exactly the format R needs. 

Here is the structure of the example shapefile folder:

```{r}
list.files('constit')
```

## Loading the data

First lets set up the names of our files so we can read them in. I'm using the filenames from my example dataset.

```{r}
fileNameShapefile <- 'constit/constituencies.shp' 

fileNameCsv <- 'ukip_candidate_seats.csv'
```

Notice <em>fileNameShapefile</em> has the name of the folder all the required parts of the shapefile are in, and then the direction to the .shp file within it.

The csv name is just the name of the csv with the extension.

Now lets load those. Reading the csv will probably be familiar.

```{r}
csv <- read.csv(fileNameCsv,stringsAsFactors = F)
```

Reading the shapefile is a little more complicated. the readOGR function requires two arguments: the dsn (where to find the file ie. <em>fileNameShapefile</em>) and the layer which is the shapefile's name without the folder at the start and the .shp on the end. 
Rather than create a new variable for that, we can write a simple regex to get the required name:

```{r}
fileNameShapefile %>% gsub('.+\\/|\\..+','',.)
```
 We can add that regex straight into the function. So all together it will look like this:

```{r}
mapData <- readOGR(dsn = fileNameShapefile,
                   layer = fileNameShapefile %>%
                     gsub('.+\\/|\\..+','',.))

```

Now we need to make sure that we have those matching columns, so lets have a look at the first five rows of our csv and the first five of our shapefile. 

This way we can work out which columns to match.

```{r}
csv %>% head(5)
```
PCON15CD (parliamentray constituency code for the year 2015) looks like the best bet here: names are more likely to have differences is spelling, case, &/and hyphens, spacing etc. Always go with codes if you can. 

Let's check to see if there is a corresponding CODE column in our shapefile. The shapefile has been read in as a spatialPolygonsDataframe. Don't worry too much about what that means; we don't have to manipulate it much here. But it means we have to access data a bit differently.

The data is stored in slot, conveniently called <em>data</em>. We access slots using @, like so:

```{r}
mapData@data %>%  head(3)
```

This all looks a bit messy, but we can see there is a column called CODE. If we select just the columns with code in,
it should look a bit cleaner.

```{r}
mapData@data %>% select(matches('CODE')) %>% head(3)
```

Great, we have a match! The only problem is the columns have different names. Let's change the name of the column in the csv, as that's easier.

```{r}
csv <- csv %>% 
  rename(CODE=PCON15CD)

names(csv)
```
Great! Now we're ready to merge the two, using our matching CODE columns.

The merge function is very simple. It takes two main arguments: the dataframe you want to merge with and the name of the column by which we want to join.
 
```{r}
mergedData <- mapData %>% 
  merge(csv,by='CODE')

mergedData %>% names

#mergedData %>% head(5) 
```

When you do this yourself, you'll want to make sure you don't have a load of NA values in your joined columns. There are some simple ways to spotcheck the data to make sure it has worked. I won't run them all in this tutorial as that would take up most of the page, but ill add a couple here:

```{r}
#mergedData %>% head(5) 

mergedData %>% as.data.frame %>%  filter(!is.na(isUkip)) %>% nrow 

# SUCCESS the number of rows with a Ukip 1 matches the number of rows in the csv

```
###SUCCESS

Now we have a merged dataset that looks like what we want to visualise in QGis. The only thing that's left is to save it as a shapefile, so we can put it straight into QGis.

We use the writeOGR() function, which takes a spatial data object (mergedData in our case), a name, a layername (keep that the same as your filename) and a driver - we want an ESRI Shapefile.


```{r}

newMapName <- 'newMap'

writeOGR(obj=mergedData, dsn=newMapName, layer=newMapName, driver="ESRI Shapefile")

```

One last point - if you save your shapefile and decide you want to change it, you'll need to delete your new folder first as it won't automatically overwrite (unlike csvs) and will error.

Hopefully that all makes sense - any questions hit me up!

### Packages used

<br>dplyr: formatting and cleaning data</br>
<br>rgdal: reading shapefiles and other spatial data</br>
<br>ggplot2: visualising data</br>
<br>rgeos: spatial data manipulation/extraction</br>




